Association between SARS-CoV-2 and metagenomic content of samples from the Huanan Seafood Market

### 2023-07-02 paper/reviewer_response.tex

Nonetheless, I think a similar analysis could have been performed using Kraken2, and I mostly took the approach here to ensure as much comparability as possible to the report by Crits-Christoph et al.

I have added text to the portion of the manuscript that explains the methods used for this part of the analysis explaining the points above.
}

3. One specific issue with the minimap2 mapping to a concatenome is how "non-uniquely" mapping reads are dealt with. "Methods" mention something about using only primary maps, but more discussion is warranted. Otherwise, reads which could be similarly diagnostic of multiple species (assuming there are some), would simply be assigned to the first genome in the concatenated string. Maybe this is not an issue here, but should be discussed.

\response{
This is a great question, because as the reviewer notes the number of aligned reads will depend on how non-uniquely mapping reads are handled.
Briefly, these are reads that map to two different species in the reference equally well.
This happens with some reads, although the rate at which it occurs is reduced by use of only mitochondrial genomes in the reference (which are much shorter than nuclear genomes) and the filtering of the reference set to remove nearly identical mitochondrial genomes.

The current pipeline only counts reads that map better to one species than any other species.
As the reviewer notes, if we simply took all primary alignments that would not be sufficient to guarantee consideration of only uniquely best mapping reads, because if a read maps equally well to two sequences (ie, species) in the reference then \texttt{minimap2} (the aligner used here) will randomly assign the primary mapping to one of the two.
(Note \texttt{minimap2} chooses the primary alignment randomly among equally good hits, not by taking the first of the concatenated genome.)
So primary aligned reads are \emph{not} guaranteed to be reads that have uniquely best mapping locations, and so could match equally well to multiple species.

However, as described briefly in the Methods of the original version of my paper, my analysis only counts primary mappings that \emph{also have a mapping quality of at least 4}.
The mapping quality is an indicator of how confidently a read maps to a specific location in the reference versus all other locations, and read will only have a mapping quality $>$0 if the primary mapping is better than all other mappings.
Therefore, the combined requirement that reads be primary and have a mapping quality greater than zero has the effect of only counting reads that uniquely map best to just one species.

I have elaborated to better explain this point in the relevant section of the Methods, as suggested by the reviewer.

Note also that in the \texttt{config.yaml} file for the pipeline, you can change the minimum mapping quality from 4 to 0 and re-run the pipeline.
If you do that, non-uniquely mapping reads are assigned randomly to one of the best-mapping reference species, which has the net effect (averaged over many reads) of giving a fractional score for these reads (eg, 0.5 for a read that maps to two species).
If you re-run the pipeline with this option, you will see that although the quantitative results shift some, the qualitative conclusions are largely unchanged.
}

### 2023-07-02 paper/paper.tex
Reviewer 2 point 3: mapping quality ensures uniquely mapped only

Briefly, the deep sequencing data were pre-processed to retain only high-quality reads.
These reads were aligned to the SARS-CoV-2 genome and a representative set of chordate mitochondrial genomes.
%The set of mitochondrial genomes was assembled by downloading all NCBI RefSeq genomes, filtering for those from chordates, and removing genomes that were highly similar to other retained genomes (within a Mash distance of $<$0.07).
This set included all the mammalian mitochondrial genomes from all species for which alignments were reported in \citet{crits2023genetic}.
The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table~\ref{tab:mito_retained}.
The counts of reads mapping to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.

@@

The sequencing reads were then aligned to this genome using \texttt{minimap2}~\citep{li2018minimap2} in the \texttt{sr} (short-read) mode.
The resulting alignments were filtered to only retain primary alignments with mapping quality of at least 4.

>>>>

Briefly, the deep sequencing data were pre-processed to retain only high-quality reads.
These reads were aligned to the SARS-CoV-2 genome and a representative set of chordate mitochondrial genomes.
%The set of mitochondrial genomes was assembled by downloading all NCBI RefSeq genomes, filtering for those from chordates, and removing genomes that were highly similar to other retained genomes (within a Mash distance of $<$0.07).
This set included all the mammalian mitochondrial genomes from all species for which alignments were reported in \citet{crits2023genetic}.
The accessions for the 4,170 chordate mitochondrial genomes that formed this set are listed in Table~\ref{tab:mito_retained}.
The counts of reads mapping uniquely to each mitochondrial genome for each sequencing run and sample are provided in Tables~\ref{tab:mito_composition_by_run} and \ref{tab:mito_composition_by_sample}, respectively.

@@

The sequencing reads were then aligned to this genome using \texttt{minimap2}~\citep{li2018minimap2} in the \texttt{sr} (short-read) mode.
The resulting alignments were filtered to only retain primary alignments with mapping quality of at least 4.
Note that one effect of counting only primary alignments with mapping quality $>$0 (4 in this case) is to only count reads that map uniquely---that is, they map better to the mitochondrial genome of one species than all other species in the reference.
This is because although \texttt{minimap2} assigns the primary mapping randomly  if there are multiple sequences in the reference with equally good mappings, a read will only have a mapping quality that exceeds zero if it maps to one location better than all other locations.
